---
title: "plots_inversions.Rmd"
author: "Andrea Estandia"
date: "17/10/2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
params:
  beagle: chr4A_1257754-chr4A_1261375.beagle
---
  
```{r setup, echo=FALSE}
# Knitr settings: 
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
options(scipen = 999)
```

```{r, echo=FALSE}
source("./src/0.0_island_rule_source.R")
```

Unzip beagle.gz window file if unzipped file does not exist already
```{r}
#if selecting individual windows
beagle_file <- file.path(
  reports_path,
    "localPCA",
    "unpruned",
    "outliers",
    "beagle_by_window",
    "max2_merge10_filter5",
    params$beagle)

if (!file.exists(beagle_file)) {
gunzip(paste0(beagle_file, ".gz"), remove=FALSE)
}
```

Read beagle file, remove allele columns, rename columns and remove extra 
numbers after sample_name
```{r}
outlier <- 
  read.table(
    file.path(
      reports_path,
      "localPCA",
      "unpruned",
      "outliers",
      "beagle_by_window",
      "max2_merge10_filter5",
      params$beagle
    ),
    header = T
  ) %>%
  select(-c("allele1", "allele2")) %>%
  as_tibble() %>%
  pivot_longer(starts_with("Ind")) %>%
  rename(sample_name = name) %>%
  rename(genotype = value) %>%
  mutate(sample_name = str_remove(sample_name, "\\.[^.]*$"))

```

Read phenotype file with body size (PC1), pop info and genetic sex
```{r}
pheno <-
  read_tsv(file.path(data_path,
                     "wgs",
                     "lists",
                     "body_pc1.tsv"),
           col_names = FALSE) %>%
  rename(sample_name = X1) %>%
  rename(body_PC1 = X2)

pop_info <- 
  read_csv(file.path(data_path,
                     "phenotypes",
                     "phenotypes+SMC2+AE.csv")) %>%
  rename(x = sample_name) %>%
  rename(sample_name = id) %>%
  select(sample_name, pop, region, latitude, longitude)

geneticsex <-
  read_csv(file.path(data_path, 
                     paste0("phenotypes/geneticsex.csv"))) %>%
  rename(id = Blood_Number)
```

Generate a vector repeating the genotypes AA, AB, and BB
Paste vector to main beagle dataframe
```{r}
geno <-
  rep(c("AA", "AB", "BB"),
      times = length(outlier$sample_name) / 3)

df <- 
  cbind(outlier, geno)

df <- df %>% 
  filter(genotype>0.34) %>% 
  group_by(sample_name, marker) %>% 
  top_n(1, genotype)
```

Merge all datasets and separate column marker into chr and position
```{r}
df2 <- 
  df %>% 
  left_join(pop_info, by = "sample_name") %>% 
  left_join(pheno, by = "sample_name") %>% 
  separate(marker,into = c("chr", "position")) %>% 
  mutate(position=as.numeric(position)) %>% 
  unite("marker", chr:position, remove=FALSE)
```

Read covariance matrix for the window and population labels to give it row names
```{r}
label <- 
  read.table(file.path(data_path,"localPCA/pop_label"))

cov_mat <-
  as.matrix(read.table(file.path(
    reports_path,
      "localPCA",
      "unpruned",
      "outliers",
      "cov_by_window",
      "max2_merge10_filter5",
      paste0(gsub("\\..*","",params$beagle), ".cov")
  )))

```

Decompose covariance matrix into its eigenvalues 
```{r}
#Do MDS on cov matrix
mds.cor <- (1 - cov_mat) %>%
  cmdscale(k=3, eig = TRUE)

colnames(mds.cor$points) <- c("Dim.1", "Dim.2", "Dim.3")
rownames(mds.cor$points) <-
  label$V3

#Do PCA on cov matrix
pca<-eigen(cov_mat)

pca.mat <-
  as.matrix(pca$vectors %*% (diag(pca$values))^0.5)

nPC <-
  dim(pca$vectors)[2]

col_PC <- 
  vector(length=nPC)

for (i in 1 : nPC) {col_PC[i] <-
  paste0("PC",i)}

#add column names
colnames(pca.mat) <-
  c(col_PC)

#add row names
rownames(pca.mat) <-
  label$V3

for (x in 1:4) {
  nam <- 
    as.character(paste0("var",x))
    assign(nam, round(pca$values[x]*100/sum(pca$values[pca$values>=0]),2))
}

kmeans_res<-
  kmeans(as.matrix(mds.cor$points[,1]), 
                   c(min(mds.cor$points[,1]),
                     mean(mds.cor$points[,1]),
                     max(mds.cor$points[,1])))
k_ss<-
  round(kmeans_res$betweenss/kmeans_res$totss,3)

k <- as.data.frame(kmeans_res$cluster)
colnames(k) <- "k"

pca.mat <- 
  as.data.frame(pca.mat)

pca.mat$pop <- 
  label$V3

pca.mat$sample_name <- 
  label$V1

pca.out <- 
  pca.mat[,c(1:4)]

clusters <- 
  cbind(label, pca.out) %>%
  select(-V3, -V4) %>% 
  cbind(mds.cor$points) %>% 
  cbind(k) %>% 
  rename(id=V1) %>% 
  rename(sample_name=V2) %>% 
  rename(subregion=V5) %>%
  left_join(geneticsex, by="id") %>% 
  left_join(df2, by="sample_name")
```

Plot PC2 vs PC1 and visually determine how many clusters
```{r}
mds_plot <- clusters %>% 
  filter(GeneticSex!="NA") %>% 
  distinct(id, .keep_all = T) %>% 
  ggplot(aes(x=PC1, y=PC2, col=as.factor(GeneticSex)))+
  geom_point()+
  theme_minimal() +
  scale_color_manual(values = c("#264653", "#e76f51"))+
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(hjust=1, size=text_size),
    axis.text.y = element_text(size = text_size),
    axis.title = element_text(size = text_size),
    legend.position = "right",
    legend.text = element_text(size=11),
    legend.title = element_text())+
  labs(x="\nMDS1", y="MDS2\n")+
       #, 
       #subtitle="Multidimensional Scaling\n")+
  guides(color=guide_legend(title="Cluster"))

ggsave(
  mds_plot,
  filename = file.path(reports_path, 
                       "plots", 
                       "inversions",
                       paste0("neosex_chr.pdf")),
  device = "pdf",width = 4.5, height=4
)
```