---
title: "plots_inversions_4rendering.Rmd"
author: "Andrea Estandia"
date: "17/10/2022"
output:
  pdf_document: default
editor_options:
  chunk_output_type: inline
params:
  beagle: beagle_list[i]
---
  
```{r setup, echo=FALSE}
# Knitr settings: 
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
options(scipen = 999)
```

```{r, echo=FALSE}
source("./src/0.0_island_rule_source.R")
```

Unzip beagle.gz window file if unzipped file does not exist already
```{r}
#if selecting individual windows
beagle_file <- file.path(
  reports_path,
    "localPCA",
    "beagle_by_window",
    params$chr,
    params$beagle)

#if selecting a set of combined outlier windows 
beagle_file <- file.path(
  reports_path,
    "localPCA",
    "outlier_mds",
    "beagle",
    params$beagle)

if (!file.exists(beagle_file)) {
gunzip(paste0(beagle_file, ".gz"), remove=FALSE)
}
```

Read beagle file, remove allele columns, rename columns and remove extra 
numbers after sample_name
```{r}
outlier <- 
  read.table(file.path(reports_path, 
                       "localPCA", 
                        "outlier_mds",
                        "beagle",
                       params$beagle), header=T) %>% 
  select(-c("allele1", "allele2")) %>% 
  as_tibble() %>% 
  pivot_longer(starts_with("Ind")) %>% 
  rename(sample_name=name) %>%
  rename(genotype=value) %>% 
  mutate(sample_name=str_remove(sample_name, "\\.[^.]*$"))

```

Read phenotype file with body size (PC1), pop info and genetic sex
```{r}
pheno <-
  read_tsv(file.path(data_path,
                     "wgs",
                     "lists",
                     "body_PC1.tsv"),
           col_names = FALSE) %>%
  rename(sample_name = X1) %>%
  rename(body_PC1 = X2)

pop_info <- 
  read_csv(file.path(data_path,
                     "phenotypes",
                     "phenotypes+SMC2+AE.csv")) %>%
  rename(x = sample_name) %>%
  rename(sample_name = id) %>%
  select(sample_name, pop, region, latitude, longitude)

geneticsex <-
  read_csv(file.path(data_path, 
                     paste0("phenotypes/geneticsex.csv"))) %>%
  rename(sample_name = Blood_Number)
```

Generate a vector repeating the genotypes AA, AB, and BB
Paste vector to main beagle dataframe
```{r}
geno <-
  rep(c("AA", "AB", "BB"),
      times = length(outlier$sample_name) / 3)

df <- 
  cbind(outlier, geno)

df <- df %>% 
  filter(genotype>0.34) %>% 
  group_by(sample_name, marker) %>% 
  top_n(1, genotype)
```

Merge all datasets and separate column marker into chr and position
```{r}
df2 <- 
  df %>% 
  left_join(pop_info, by = "sample_name") %>% 
  left_join(pheno, by = "sample_name") %>% 
  separate(marker,into = c("chr", "position")) %>% 
  mutate(position=as.numeric(position)) %>% 
  unite("marker", chr:position, remove=FALSE)
```

Read covariance matrix for the window and population labels to give it row names
```{r}
label <- 
  read.table(file.path(data_path,"localPCA/pop_label"))

cov_mat <-
  as.matrix(read.table(file.path(
    reports_path,
      "localPCA",
      "outlier_mds",
      "cov",
      paste0(gsub("\\..*","",params$beagle), ".cov")
  )))

```

Decompose covariance matrix into its eigenvalues 
```{r}
#Do MDS on cov matrix
mds.cor <- (1 - cov_mat) %>%
  cmdscale(k=3, eig = TRUE)

colnames(mds.cor$points) <- c("Dim.1", "Dim.2", "Dim.3")
rownames(mds.cor$points) <-
  label$V3

#Do PCA on cov matrix
pca<-eigen(cov_mat)

pca.mat <-
  as.matrix(pca$vectors %*% (diag(pca$values))^0.5)

nPC <-
  dim(pca$vectors)[2]

col_PC <- 
  vector(length=nPC)

for (i in 1 : nPC) {col_PC[i] <-
  paste0("PC",i)}

#add column names
colnames(pca.mat) <-
  c(col_PC)

#add row names
rownames(pca.mat) <-
  label$V3

for (x in 1:4) {
  nam <- 
    as.character(paste0("var",x))
    assign(nam, round(pca$values[x]*100/sum(pca$values[pca$values>=0]),2))
}

kmeans_res<-
  kmeans(as.matrix(mds.cor$points[,1]), 
                   c(min(mds.cor$points[,1]), 
                     median(mds.cor$points[,1]), 
                     max(mds.cor$points[,1])))
k_ss<-
  round(kmeans_res$betweenss/kmeans_res$totss,3)

k <- as.data.frame(kmeans_res$cluster)
colnames(k) <- "k"

pca.mat <- 
  as.data.frame(pca.mat)

pca.mat$pop <- 
  label$V3

pca.mat$sample_name <- 
  label$V1

pca.out <- 
  pca.mat[,c(1:4)]

clusters <- 
  cbind(label, pca.out) %>%
  select(-V3, -V4) %>% 
  cbind(mds.cor$points) %>% 
  cbind(k) %>% 
  rename(id=V1) %>% 
  rename(sample_name=V2) %>% 
  rename(subregion=V5) %>%
  left_join(geneticsex) %>% 
  left_join(df2, by="sample_name")
```

Plot PC2 vs PC1 and visually determine how many clusters
```{r}
mds_plot <- clusters %>% 
  distinct(id, .keep_all = T) %>% 
  ggplot(aes(y=Dim.2, x=Dim.1, col=as.factor(k)))+
  geom_point()+
  theme_minimal() +
  scale_color_manual(values = c("#264653", "#2a9d8f", "#e9c46a"))+
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(hjust=1, size=text_size),
    axis.text.y = element_text(size = text_size),
    axis.title = element_text(size = text_size),
    legend.position = "right",
    legend.text = element_text(size=11),
    legend.title = element_text())+
  labs(x="\nMDS1", y="MDS2\n", 
       title=paste0("Outlier window in ", gsub("\\_.*","",params$beagle), "\n"), 
       subtitle="Multidimensional Scaling\n")+
  guides(color=guide_legend(title="Cluster"))

pca_plot <- clusters %>% 
  distinct(id, .keep_all = T) %>% 
  ggplot(aes(y=PC2, x=PC1, col=as.factor(k)))+
  geom_point()+
  theme_minimal() +
  scale_color_manual(values = c("#264653", "#2a9d8f", "#e9c46a"))+
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    axis.text.x = element_text(hjust=1, size=text_size),
    axis.text.y = element_text(size = text_size),
    axis.title = element_text(size = text_size),
    legend.position = "right",
    legend.text = element_text(size=11),
    legend.title = element_text())+
  labs(x="\nPC1", y="PC2\n",
       subtitle="PCA\n")+
  guides(color=guide_legend(title="Cluster"))

plot_mds_pca <- 
  mds_plot+pca_plot+plot_layout(guides = "collect")

plot_mds_pca

ggsave(
  plot_mds_pca,
  filename = file.path(reports_path, 
                       "plots", 
                       paste0("mds_vs_pca_",params$beagle,".pdf")),
  device = "pdf",width = 10, height=5
)
```

<!-- If 2D is not clear plot in 3D and explore groupings -->
<!-- ```{r} -->
<!-- df <-  -->
<!--   clusters %>%  -->
<!--   distinct(sample_name, .keep_all = T) -->

<!-- plot_ly(x=df$PC1,  -->
<!--         y=df$PC2,  -->
<!--         z=df$PC3,  -->
<!--         type="scatter3d",  -->
<!--         mode="markers",  -->
<!--         color=df$k) -->
<!-- ``` -->

<!-- Select positions of window to set as plot subtitle and make three plots: -->
<!-- 1. PC2 vs PC1 coloured by group of choice -->
<!-- 2. Tiles by population -->
<!-- 3. Tiles by region -->

<!-- Save plots  -->
<!-- ```{r} -->
<!-- positions <-  -->
<!-- as.vector(rbind(min(clusters$position), max(clusters$position))) -->

<!-- palette_Dark2 <-  -->
<!--   colorRampPalette(brewer.pal(14, "Set2")) -->

<!-- pal <-  -->
<!--   palette_Dark2(length(unique(clusters$pop))) -->

<!-- plot1 <-  -->
<!--   clusters %>% -->
<!--   distinct(id, .keep_all = T) %>%  -->
<!--   ggplot(aes(x = Dim.1, y = Dim.2, col=as.factor(k))) + -->
<!--   geom_point() + -->
<!--   scale_color_manual(values = c("#264653", -->
<!--                                 "#2a9d8f",  -->
<!--                                 "#e9c46a"))+ -->
<!--   theme_minimal() + -->
<!--   theme( -->
<!--     panel.border = element_blank(), -->
<!--     panel.grid.major = element_blank(), -->
<!--     panel.grid.minor = element_blank(), -->
<!--     axis.text.x = element_text( -->
<!--       hjust = 1, -->
<!--       size = text_size -->
<!--     ), -->
<!--     axis.text.y = element_text(size = text_size), -->
<!--     axis.title = element_text(size = text_size), -->
<!--     legend.position = "bottom", -->
<!--     legend.text = element_text(size = 11), -->
<!--     legend.title = element_text() -->
<!--   ) + -->
<!--   labs( -->
<!--     x = "\nMDS1", -->
<!--     y = "MDS2\n", -->
<!--     title = paste0("Chr30: ", positions[1], "-", positions[2]))+  -->
<!--   guides(col=guide_legend(title="")) -->


<!-- plot2 <- clusters %>% -->
<!--   arrange(position) %>% -->
<!--   ggplot(aes( -->
<!--     y = as.factor(k), -->
<!--     x = fct_reorder(marker, position), -->
<!--     fill = geno, -->
<!--     col = geno -->
<!--   )) + -->
<!--   geom_tile() + -->
<!--   scale_fill_manual(values = c("#264653", -->
<!--                                 "#2a9d8f",  -->
<!--                                 "#e9c46a"))+ -->
<!--   scale_color_manual(values = c("#264653", -->
<!--                                 "#2a9d8f",  -->
<!--                                 "#e9c46a"))+ -->
<!--   theme_minimal() + -->
<!--   theme( -->
<!--     axis.text.x = element_blank(), -->
<!--     axis.text.y = element_blank(), -->
<!--     axis.title = element_text(size = text_size), -->
<!--     legend.title = element_text(), -->
<!--     legend.position = "bottom", -->
<!--     panel.grid.major = element_blank(), -->
<!--     strip.text.y.right = element_text(angle = 0), -->
<!--     strip.text.y.left = element_blank() -->
<!--   ) + -->
<!--   labs( -->
<!--     x = "\nSNPs", -->
<!--     y = "Cluster\n", -->
<!--     title = paste0("Inversion in ",params$chr), -->
<!--     subtitle = paste0("location: ", positions[1], "-", positions[2]) -->
<!--   ) + -->
<!--   facet_grid(rows = vars(factor(k, levels = c( -->
<!--     "1", "2" -->
<!--   )),region)) -->

<!-- plot3 <- clusters %>% -->
<!--   filter(region!="NA") %>%  -->
<!--   arrange(position) %>% -->
<!--   ggplot(aes( -->
<!--     y = as.factor(k), -->
<!--     x = fct_reorder(marker, position), -->
<!--     fill = geno, -->
<!--     col = geno -->
<!--   )) + -->
<!--   geom_tile() + -->
<!--   scale_fill_manual(values = c("#cc4b3f", "#279ae6", "#e6bf27")) + -->
<!--   scale_color_manual(values = c("#cc4b3f", "#279ae6", "#e6bf27")) + -->
<!--   theme_minimal() + -->
<!--   theme( -->
<!--     axis.text.x = element_blank(), -->
<!--     #axis.text.y = element_blank(), -->
<!--     axis.title = element_text(size = text_size), -->
<!--     legend.title = element_text(), -->
<!--     legend.position = "bottom", -->
<!--     panel.grid.major = element_blank(), -->
<!--     # strip.text.y.right = element_text(angle = 0), -->
<!--     #strip.text.y.left = element_blank() -->
<!--   ) + -->
<!--   labs( -->
<!--     x = "\nSNPs", -->
<!--     y = "Cluster\n", -->
<!--     title = paste0("Inversion in ",params$chr), -->
<!--     subtitle = paste0("location: ", positions[1], "-", positions[2]) -->
<!--   )  + -->
<!--   facet_grid(rows = .~region) -->


<!-- common_plot <- -->
<!--   (plot1+plot3)/plot2+plot_layout(nrow=2) -->

<!-- common_plot -->

<!-- ggsave( -->
<!--   common_plot, -->
<!--   filename = file.path(reports_path,  -->
<!--                        "localPCA",  -->
<!--                        "plots",  -->
<!--                        "windows", -->
<!--                        paste0(params$beagle,".pdf")), -->
<!--   device = "pdf",width = 15, height=8 -->
<!-- ) -->

<!-- clusters %>% -->
<!--   filter(genotype>0.34) %>%  -->
<!--   group_by(marker, geno, region) %>%  -->
<!--   summarise(n=n()) %>%  -->
<!--   ggplot(aes(x=marker,y=n,fill=geno))+ -->
<!--   geom_bar(stat="identity") -->

<!-- ``` -->

```{r}
df <- 
  clusters %>% 
  group_by(pop, geno, k, region) %>% 
  summarise(n=n())

regions <-
  df %>% 
  distinct(region) %>% 
  drop_na() 

for (i in as.vector(regions$region)) {
  #sum all genotypes by region
  total_geno <- df %>%
    group_by(region) %>%
    drop_na() %>%
    summarise(total_n = sum(n))
  #sum all genotypes containing allele B
  total_b <- df %>%
    filter(grepl("B", geno)) %>%
    group_by(region) %>%
    drop_na() %>%
    summarise(total_b = sum(n))
  #merge dataset
  summary_alleles_b <-
    full_join(total_geno, total_b)
  
}

summary_alleles_b$percentage_b <- summary_alleles_b[,3]/summary_alleles_b[,2]
summary_alleles_b

regions <-
  df %>% 
  distinct(pop) %>% 
  drop_na() 

for (i in as.vector(regions$pop)) {
  #sum all genotypes by region
  total_geno <- df %>%
    group_by(pop) %>%
    drop_na() %>%
    summarise(total_n = sum(n))
  #sum all genotypes containing allele B
  total_cluster1 <- df %>%
    filter(grepl("1", as.factor(k))) %>%
    group_by(pop) %>%
    drop_na() %>%
    summarise(total_cluster1 = sum(n))
  total_cluster2 <- df %>%
    filter(grepl("2", as.factor(k))) %>%
    group_by(pop) %>%
    drop_na() %>%
    summarise(total_cluster2 = sum(n))
  total_cluster3 <- df %>%
    filter(grepl("3", as.factor(k))) %>%
    group_by(pop) %>%
    drop_na() %>%
    summarise(total_cluster3 = sum(n))
  #merge dataset
  summary_alleles_cluster <-
    full_join(total_geno, total_cluster1) %>% 
    full_join(total_cluster2) %>%  
    full_join(total_cluster3)
  
}

summary_alleles_cluster$percentage_cluster1 <- 
  summary_alleles_cluster[,3]/summary_alleles_cluster[,2]

summary_alleles_cluster$percentage_cluster2 <- 
  summary_alleles_cluster[,4]/summary_alleles_cluster[,2]

summary_alleles_cluster$percentage_cluster3 <- 
  summary_alleles_cluster[,5]/summary_alleles_cluster[,2]

summary_alleles_cluster

write_csv(df_map,
          file.path(reports_path, 
                   "localPCA", 
                   "outlier_mds", 
                   paste0(params$beagle,"_summary_alleles_cluster.csv")))
```

```{r}
df_map <-
  summary_alleles_cluster %>% 
  left_join(pop_info, by="pop") %>% 
  distinct(pop, .keep_all = T) %>% 
  select(-sample_name)

df_map$percentage_cluster1 <-
  as.numeric(unlist(df_map$percentage_cluster1))
df_map$percentage_cluster2 <-
  as.numeric(unlist(df_map$percentage_cluster2))
df_map$percentage_cluster3 <-
  as.numeric(unlist(df_map$percentage_cluster3))

df_map[is.na(df_map)] <- 0

oceania <-
  rnaturalearth::ne_countries(scale = "large",
               returnclass = "sf",
               continent = "oceania")

anzo <- ggplot(data = oceania) +
    geom_sf(fill = "darkgrey", color = NA) +
    coord_sf(xlim = c(140, 187),
             ylim = c(-50,-12),
             expand = FALSE) +
    theme(
      text=element_text(),
      panel.background = element_rect(fill = background_colour),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks = element_blank(),
      # legend.position = "none",
      plot.margin = margin(0.2, 0.2, 0.2, 0.2, "cm"))+
    geom_scatterpie(aes(x=longitude+0.5, y=latitude+0.5, group = pop, r = 1), 
                    data = df_map, cols=c("percentage_cluster1", 
                                          "percentage_cluster2", 
                                          "percentage_cluster3"), color=NA) + 
    scale_fill_manual(values=c("#4483a2", 
                               "#e38440",
                               "#d9c31e"))

ggplot(data = oceania) +
    geom_sf(fill = "darkgrey", color = NA) +
    coord_sf(xlim = c(-180, -173),
             ylim = c(-47,-40),
             expand = FALSE) +
    theme(
      text=element_text(),
      panel.background = element_rect(fill = background_colour),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks = element_blank(),
      # legend.position = "none",
      plot.margin = margin(0.2, 0.2, 0.2, 0.2, "cm"))+
    geom_scatterpie(aes(x=longitude+0.5, y=latitude+0.5, group = pop, r = 1), 
                    data = df_map, cols=c("percentage_cluster1", 
                                          "percentage_cluster2", 
                                          "percentage_cluster3"), color=NA) + 
    scale_fill_manual(values=c("#4483a2", 
                               "#e38440",
                               "#d9c31e"))

ggplot(data = oceania) +
    geom_sf(fill = "darkgrey", color = NA) +
    coord_sf(xlim = c(-155, -143),
             ylim = c(-20,-14),
             expand = FALSE) +
    theme(
      text=element_text(),
      panel.background = element_rect(fill = background_colour),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks = element_blank(),
      # legend.position = "none",
      plot.margin = margin(0.2, 0.2, 0.2, 0.2, "cm"))+
    geom_scatterpie(aes(x=longitude+0.5, y=latitude+0.5, group = pop, r = 1), 
                    data = df_map, cols=c("percentage_cluster1", 
                                          "percentage_cluster2", 
                                          "percentage_cluster3"), color=NA) + 
    scale_fill_manual(values=c("#4483a2", 
                               "#e38440",
                               "#d9c31e"))

van_sm <- ggplot(data = oceania) +
    geom_sf(fill = "darkgrey", color = NA) +
    coord_sf(xlim = c(160, 172),
             ylim = c(-25,-12),
             expand = FALSE) +
    theme(
      text=element_text(),
      panel.background = element_rect(fill = background_colour),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.text.y = element_blank(),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.ticks = element_blank(),
      # legend.position = "none",
      plot.margin = margin(0.2, 0.2, 0.2, 0.2, "cm"))+
    geom_scatterpie(aes(x=longitude+0.5, y=latitude+0.5, group = pop, r = 0.3), 
                    data = df_map, cols=c("percentage_cluster1", 
                                          "percentage_cluster2", 
                                          "percentage_cluster3"), color=NA) + 
    scale_fill_manual(values=c("#4483a2", 
                               "#e38440",
                               "#d9c31e"))

chartpie <- 
  anzo+van_sm+plot_layout(guides="collect")

chartpie

ggsave(
  chartpie,
  filename = file.path(reports_path, 
                       "plots", 
                       paste0("chartpie_",params$beagle,".pdf")),
  device = "pdf",width = 10, height=5
)
```

